What We Eval in the Shadows

### 2021-04-14 analysis/side-effects.Rmd
Side effects

library(DT)
library(digest)
library(fs)
library(fst)
library(purrr)
library(tidyverse)

@@

create_tags(path(TAGS_DIR, "side-effects.tex"), prefix = "", default = TRUE)
```

```{r load data}
SE <- read_fst(PACKAGE_SIDE_EFFECTS_FILE) %>% as_tibble
```

```{r fix withr::defer callsite}

```{r fix S3 methods}

>>>>

library(DT)
library(digest)
library(fs)
library(fst)
library(purrr)
library(tidyverse)
library(xtable)

@@

create_tags(path(TAGS_DIR, "side-effects.tex"), prefix = "SE", default = TRUE)
```

```{r load data}
SE_all <- read_fst(PACKAGE_SIDE_EFFECTS_FILE) %>% 
  as_tibble %>%
  filter(transitive==0)
```

```{r missing calls}
SE_missing_eval_call <- filter(SE_all, is.na(eval_function))
SE_all <- filter(SE_all, !is.na(eval_function))
```

```{r deal with empty srcrefs}
SE <- mutate(SE_all, eval_call_srcref=case_when(
  eval_call_srcref == "???" & expr_expression == "handler$expr" & envir_expression == "handler$envir" ~ "::withr::execute_handlers::1",
  eval_call_srcref == "???" & expr_expression == "quote({\n    old.options <- options(dplyr.summarise.inform = FALSE)\n    o" ~ "::dplyr::tally::1",
  eval_call_srcref == "???" & expr_expression == "quote({\n    na.action <- attr(mf, \"na.action\")\n    why_omit <- attr(na.a" ~ "::estimatr::cleam_mode_data::1",
  TRUE ~ eval_call_srcref
))

SE_missing_sites <- filter(SE, eval_call_srcref=="???")
SE <- filter(SE, eval_call_srcref!="???")
```

```{r remove Random.seed}
SE_random_seed <- filter(SE, variable == ".Random.seed", !str_detect(expr_resolved, "set.seed"))
SE <- filter(SE, variable != ".Random.seed" | str_detect(expr_resolved, "set.seed"))
```

```{r remove internal}
SE_internal <- filter(SE, str_starts(source, fixed("internal:")))
SE <- filter(SE, !str_starts(source, fixed("internal:")))
```

```{r load static evals}
ES <- read_csv(PACKAGE_EVALS_STATIC_FILE) %>% rename(eval_call_srcref=srcref)
```

```{r load summarize data}
S <- read_fst(PACKAGE_SUM_CALLS_FILE) %>% as_tibble
```

```{r}
CRAN <- read_fst(PACKAGE_CORPUS_FILE) %>% as_tibble
PKG <- filter(CRAN, in_corpus) 
```

## Overview

```{r}
SE_all_N <- sum(SE_all$N)
SE_all_sites <- count(SE_all, eval_call_srcref)
SE_all_calls <- count(SE_all, cid)

SE_internal_N <- sum(SE_internal$N)
SE_internal_sites <- count(SE_internal, eval_call_srcref)

SE_random_seed_N <- sum(SE_random_seed$N)
SE_random_seed_sites <- count(SE_random_seed, eval_call_srcref)

SE_N <- sum(SE$N)
SE_calls <- count(SE, cid)
SE_sites <- count(SE, eval_call_srcref)
SE_missing_N <- sum(SE_missing_sites$N)

overview_table(
  r("All", SE_all_N),
  r("All calls", SE_all_calls),
  r("All sites", SE_all_sites),

  r("Internal", SE_internal_N),
  r("Internal ratio", ratio(SE_internal_N, SE_all_N)),

  r("Internal sites", SE_internal_sites),
  r("Internal sites ratio", ratio(SE_internal_sites, SE_all_sites)),

  r("Random seed", SE_random_seed_N),
  r("Random seed ratio", ratio(SE_random_seed_N, SE_all_N)),

  r("Random seed sites", SE_random_seed_sites),
  r("Random seed sites ratio", ratio(count(SE_random_seed, eval_call_srcref), SE_all_sites)),

  r("Missing SE sites", SE_missing_N),
  r("Missing SE sites ratio", ratio(SE_missing_N, SE_N)),

  r("User", SE_N),
  r("User ratio", ratio(SE_N, SE_all_N)),

  r("User calls", SE_calls),
  r("User calls ratio", ratio(SE_calls, SE_all_calls)),

  r("User sites", SE_sites),
  r("User sites ratio", ratio(SE_sites, SE_all_sites))
)
```

## Classification

```{r functions}
SE <- mutate(
  SE, 
  class=case_when(
    source == str_c("call:", caller_function)           ~ "Local",
    str_starts(source, fixed("call:"))                  ~ "Function",
    str_detect(source, "^R6:.*R6.*") | source=="???S4"  ~ "Object",
    source == "pkg:global"                              ~ "Global",
    str_starts(source, fixed("pkg:"))                   ~ "Package",
    TRUE                                                ~ "Synthetic"
  ),
  caller=str_replace(eval_call_srcref, "::(.*)::(.*)::.*", "\\1::\\2")
)
```

```{r}
sites_classes <- 
  SE %>%
  group_by(eval_call_srcref) %>%
  summarise(n_class=n_distinct(class), classes=str_c(sort(unique(class)), collapse=", "))

cs_count(sites_classes, n_class)
cs_count(sites_classes, classes)
cs_count(SE, class)
```

```{r}
funs <- cs_count(SE, caller)
```

```{r}
sites_envir <- 
  SE %>%
  group_by(eval_call_srcref) %>% 
  summarize(
    in_envir=case_when(
      all(!in_envir) ~ 0,
      all(in_envir)  ~ 1,
      TRUE ~ 2
    )
  ) %>% 
  ungroup()
```

```{r}
sites_not_in_envir <- filter(sites_envir, in_envir==0)
sites_in_envir <- filter(sites_envir, in_envir==1)
sites_both_envir <- filter(sites_envir, in_envir==2)

overview_table(
  r("Funs nighty", filter(funs, cp <= 90)),
  r("Funs ten", filter(funs, cp > 90)),
  r("Sites not in envir", sites_not_in_envir),
  r("Sites not in envir ratio", ratio(sites_not_in_envir, sites_envir)),
  r("Sites in envir", sites_in_envir),
  r("Sites in envir ratio", ratio(sites_in_envir, sites_envir)),
  r("Sites both envir", sites_both_envir),
  r("Sites both envir ratio", ratio(sites_both_envir, sites_envir))
)
```

```{r}
top_sites_classes <- cs_count(sites_classes, classes) %>% filter(cp<=91)
```

```{r}
print(
  xtable(
    mutate(
      top_sites_classes,
      p=fmt(p, suffix="%", digit=1),
      cp=fmt(cp, suffix="%", digit=1),
    ), 
    align="llrrr",
  ),
  file=file.path(TAGS_DIR, "table-top-sites-classes-se.tex"),
  comment=FALSE, 
  include.colnames=FALSE,
  include.rownames=FALSE,
  only.contents=TRUE,
  timestamp=FALSE
)
```

## Function classification

```{r}
single_class_funs <- 
  left_join(
    mutate(ES, fun=str_c(package, fun_name)), 
    sites_classes, 
    by="eval_call_srcref"
  ) %>% 
  group_by(fun) %>% 
  filter(all(!is.na(n_class))) %>%
  summarize(n_classes=n_distinct(classes), classes=str_c(unique(classes), collapse=";"))

cs_count(single_class_funs, n_classes)
cs_count(single_class_funs, classes)
```

```{r}
single_class_funs %>%
  mutate(
    classes=if_else(str_detect(classes, "[,;]"), "\\emph{multiple}", classes)
  ) %>%
  cs_count(classes) %>%
  mutate(
    p=fmt(p, suffix="\\%", digit=1),
    o=if_else(str_starts(classes, fixed("\\")), 0L, n)
  ) %>% 
  arrange(desc(o)) %>%
  select(-cp, -o) %>%
  xtable(
    align="llrr",
  ) %>% 
  print(
    file=file.path(TAGS_DIR, "table-top-sites-classes-funs.tex"),
    comment=FALSE, 
    include.colnames=FALSE,
    include.rownames=FALSE,
    only.contents=TRUE,
    sanitize.text.function=identity,
    timestamp=FALSE
  )
```

### 2021-04-14 paper/main.tex
Side effects

@@

\subsection{Eval in R}

@@

\subsection{The environments of \eval}

@@

\todo{Filip: discuss the side-effects}

>>>>

% cf. https://tex.stackexchange.com/a/144640
\makeatletter\let\expandableinput\@@input\makeatother

@@

\subsection{Eval in R}\label{sec:eval-in-r}

@@

\subsection{The environments of \eval}\label{sec:env}

@@

The code executed by \eval can do side effects. From the compiler perspective,
we care about the observable side effects in environments---\ie variable
definitions, updates, and removals that are visible after a call to \eval
finishes. Knowing where---\ie in which environments do these side-effects
happen, can help us to determine how much of the compiler knowledge about the
program will be potentially invalidated.

The \rdyntrace contains low-level hooks capturing the life-cycle of an
environment and variables defined in it. In the tracer, we register callbacks
that record each environment creation and removal together with updates to its
bindings. From the recorded data, we ignore side effects
coming from \eval sites from unit testing frameworks.\footnote{In the corpus we
have \c{RUnit, testthat, tinytest} and \c{unitizer} unit testing frameworks.}
They run the testing code via \eval and thus the results would be biased
because of the high number of code run in these tests
(\CranRunnableCodeTestsRnd lines).

From the \packageNbrunsRnd programs we capture \SEAll side effects from
\SEAllCalls \eval calls in \SEAllSites sites. The challenge is again to remove
the accidental side-effects that are caused by the R virtual machine
implementation and are not related to the user code. For example, the
\c{.Random.seed} variable, which contains the state of the random number
generator is saved and restored from and to the global environment every time
user calls one of the base routines for getting random numbers. Removing them
leaves us with \SEUser side effects (\SEUserRatio) from \SEUserCalls \eval
calls (\SEUserCallsRatio) in \SEUserSites sites (\SEUserSitesRatio). Half of
the side-effects comes from three functions: \c{plyr::allocate\_column}
(allocates space for a new data frame column), \c{withr::execute\_handlers}
(executes deferred expressions) and \c{foreach::doSEQ} (executes an expression
on each element in a collection, possibly in parallel). \SEFunsNighty functions
are responsible for 90\% of side effects. The remaining is spread in \SEFunsTen
functions.

Most of the \eval sites (\SESitesInEnvirRatio) do side effects in the
environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}). \SESitesNotInEnvirRatio modifies other
environment and finally \SESitesBothEnvirRatio does both.
Table~\ref{tab:side-effects-targets} shows the class of the environments where
\eval side effects happen. It follows the classification described in
Section~\ref{sec:env} with two extra classes: \emph{Object} which represents a
side-effect that happens in S4 or R6 object environment, and \emph{Local} that
corresponds to a side effect in local variables of the \eval caller.

\begin{table}[h]\centering\small
  \begin{tabular}{l|r|r|r}\hline
    Environments & \#sites & \%sites & cum. \%sites \\%
    \expandableinput tag/table-top-sites-classes-se.tex
  \end{tabular}
  \caption{Target environments for \eval side-effects} \label{tab:side-effects-targets}
\end{table}

The good news is that that most \eval sites do all side effects consistently in
the same environment and that almost half of the side effects happen in either
\emph{Local} or \emph{Object} environments. These are the cases the compiler
could possibly deal with.
\todo{Finish text for Table~\ref{tab:se-funs}}

\begin{table}[h]\centering\small
  \begin{tabular}{l|r|r}\hline
    Environment & \#functions & \%functions \\%
    \expandableinput tag/table-top-sites-classes-funs.tex
  \end{tabular}
  \caption{Number of functions in which all \eval side effects happen in the same environment class} \label{tab:se-funs}
\end{table}

\mypara{Discussion}
\todo{Is there any comparison with JS?}
